---
title: "deepwikiR: Automated R Code Documentation and Exploration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{deepwikiR: Automated R Code Documentation and Exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Introduction

`deepwikiR` is an advanced R package designed to automate the documentation and exploration of R code repositories. It leverages static code analysis, the power of Large Language Models (LLMs) for generating human-like text, and the Quarto publishing framework to produce comprehensive and well-structured documentation.

This tool helps developers and teams to:

*   Understand complex R projects more quickly.
*   Maintain up-to-date documentation with minimal effort.
*   Explore codebase structures through visualizations and interactive queries.

### Key Features

*   **Automated Code Documentation:** Generates detailed descriptions for R functions and other code elements using LLMs.
*   **Call Graph Visualization:** Creates Mermaid diagrams to visualize function and module dependencies, offering insights into code structure.
*   **Multi-Project Support:** Manages and processes multiple R projects from a single configuration file, producing a master index page for easy navigation of all generated documentations.
*   **Directory-Level Summaries:** Provides LLM-generated summaries highlighting the purpose and contents of each directory within your project.
*   **Interactive Code Chat:** Allows users to ask questions about an analyzed codebase in an interactive command-line session, receiving answers from an LLM informed by the codebase's documentation.
*   **Fine-tuning Data Export:** Exports documented code elements in JSONL format, suitable for fine-tuning custom language models on your specific coding patterns and documentation style.
*   **Quarto Integration:** Outputs documentation in various formats (HTML, PDF, etc.) using the versatile Quarto framework.
*   **R Package Structure Adherence:** Built with R package best practices in mind, utilizing Roxygen2 for inline documentation.

## 2. Installation and Setup

### System Requirements

*   R (version 4.0.0 or higher recommended).
*   Quarto CLI (version 1.0.0 or higher recommended). Download from [quarto.org](https://quarto.org/docs/get-started/).
*   Git (must be installed and in the system PATH if you plan to analyze remote Git repositories).

### R Package Dependencies

`deepwikiR` relies on several R packages listed in its `DESCRIPTION` file. Key dependencies include:

*   `argparse`: For command-line argument parsing.
*   `CodeDepends`: For static R code analysis (though currently used minimally, with potential for deeper integration).
*   `git2r`: For cloning Git repositories.
*   `jsonlite`: For reading and writing JSON configuration files.
*   `quarto`: For rendering Quarto documents.
*   `stringr`: For string manipulation.
*   `tidyllm`: For interacting with LLM APIs (currently focused on OpenAI-compatible ones).

Other packages like `devtools`, `roxygen2`, and `testthat` are suggested for development and testing.

### Installation Steps

1.  **Install R and Quarto:** Follow instructions from their respective official websites.
2.  **Install `deepwikiR`:**
    Currently, `deepwikiR` is under development. To use it, you would typically clone its repository.

    ```bash
    git clone https://github.com/your_username/deepwikiR.git # Replace with actual URL
    cd deepwikiR
    ```

    Then, install dependencies and the package itself. From an R session within the cloned project's root:

    ```{r eval=FALSE}
    # Install dependencies (this might be more streamlined with pak or remotes in the future)
    # For now, ensure you have the packages listed in DESCRIPTION's Imports and Suggests.
    # Example for some key ones:
    install.packages(c("argparse", "git2r", "jsonlite", "quarto", "stringr", "tidyllm", "devtools", "roxygen2", "testthat", "CodeDepends"))

    # Build and install deepwikiR locally
    devtools::install()
    ```

3.  **Set up LLM API Key:**
    `deepwikiR` requires access to an LLM. For OpenAI models, set your API key as an environment variable. The default environment variable name is `OPENAI_API_KEY`. You can specify a different variable name in the configuration.

    ```bash
    # In your .bashrc, .zshrc, or .Renviron
    export OPENAI_API_KEY="your_openai_api_key_here"
    ```
    Remember to restart your R session or terminal for the environment variable to take effect.

## 3. Core Concepts

### Multi-Project Configuration

`deepwikiR` uses a single JSON configuration file to manage documentation generation for one or more projects. This file is typically structured with `global_settings` and a `projects` array.

*   **`global_settings`**: Defines parameters that apply to all projects unless overridden. This includes:
    *   `output_dir_root`: The main directory where all project documentation outputs will be nested.
    *   `default_llm_settings`: Default LLM configurations (provider, model, API details, etc.).
*   **`projects`**: An array where each object defines a specific project to be documented. Each project configuration can specify:
    *   `project_name`: A unique identifier for the project.
    *   `code_source`: Details about where the code is located (e.g., `local_path` or `git_repo`), language hints, and include/exclude patterns.
    *   `output_dir`: A directory name (relative to `output_dir_root`) for this project's documentation.
    *   `output_filename_base`: The base name for the main Quarto document.
    *   `llm_settings`: Project-specific LLM settings that override the global defaults.
    *   `export_finetuning_data`: Settings for exporting data for LLM fine-tuning.

Here's a snippet illustrating the structure:

```json
{
  "global_settings": {
    "output_dir_root": "output/my_docs_hub",
    "default_llm_settings": {
      "provider": "openai",
      "model": "gpt-3.5-turbo",
      "api_details": { "api_key_env_var": "OPENAI_API_KEY" }
    }
  },
  "projects": [
    {
      "project_name": "MyFirstProject",
      "code_source": { "local_path": "path/to/project1_R_code" },
      "output_dir": "ProjectOneDocs",
      "output_filename_base": "project_one_main_doc"
      // Uses default_llm_settings
    },
    {
      "project_name": "AnotherProject",
      "code_source": { "git_repo": "https://github.com/user/repo2.git" },
      "output_dir": "ProjectTwoDocs",
      "output_filename_base": "project_two_main_doc",
      "llm_settings": { "model": "gpt-4" } // Overrides default model
    }
  ]
}
```

### Documentation Generation Workflow

For each project defined in the configuration, `deepwikiR` performs these steps:

1.  **Code Acquisition:** Fetches code from the specified local path or Git repository.
2.  **Code Analysis:**
    *   Extracts R functions and their basic properties (name, signature, code block).
    *   Builds a call graph to understand dependencies between functions.
    *   Lists the content of each directory.
3.  **LLM-Powered Documentation:**
    *   **Element Documentation:** For each extracted function, an LLM generates a concise description of its purpose, inputs, and outputs.
    *   **Directory Summaries:** For each directory, an LLM generates a high-level summary of its role and contents, based on the files it contains and the documentation of elements within those files.
4.  **Quarto Rendering:** All collected information (directory summaries, call graph, function documentation) is compiled into a Quarto (`.qmd`) file and rendered into the desired output format (e.g., HTML).
5.  **Data Export (Optional):** If configured, documented elements are exported in JSONL format for LLM fine-tuning.
6.  **Analysis Data Caching:** Key analysis results (documented elements, call graph, summaries) are saved in an `.rds` file in the project's output directory. This file is used for the interactive chat mode.

### Master Index Page

When processing multiple projects defined in the configuration and a `global_settings.output_dir_root` is set, `deepwikiR` generates a master `index.html` page in that root directory. This page provides links to the individual documentation generated for each project, creating a centralized hub.

## 4. Detailed Feature Explanations

### Multi-Project Processing

As described above, you can manage documentation for several R projects through one `config.json`. When you run `deepwikiR` in `generate` mode with such a config, it will iterate through each project entry in the `projects` array, generating documentation for each according to its specific (or inherited global) settings.

### Directory-Level Summaries

A unique feature of `deepwikiR` is its ability to provide LLM-generated summaries for each directory in your project. This helps in understanding the high-level organization and purpose of different parts of your codebase.

In the generated Quarto document for a project, these summaries appear in a dedicated "Directory Overview" section, typically before the detailed function documentation. Each directory is listed with its summary.

### Interactive Chat Mode

Once a project's documentation (and its associated `_analysis_data.rds` file) has been generated, you can use the interactive chat mode to ask questions about the codebase. This feature uses an LLM, providing it with the context of your project's documented elements.

To use this mode:

*   **From an R session (if `deepwikiR` is loaded):**
    ```{r eval=FALSE}
    deepwikiR::chat_with_repo(
      analysis_data_path = "path/to/project_output_dir/your_project_analysis_data.rds",
      config_path = "path/to/your_main_config.json"
      # The config_path is needed for LLM settings
    )
    ```
*   **From the command line (recommended for CLI usage):**
    ```{bash eval=FALSE}
    # Ensure R/deepwikiR.R is executable or use Rscript path/to/deepwikiR_project_root/R/deepwikiR.R ...
    Rscript R/deepwikiR.R chat \
      --data_file path/to/project_output_dir/your_project_analysis_data.rds \
      --config_file path/to/your_main_config.json
    ```

This will start a session where you can type questions (e.g., "What does function X do?", "Which functions are related to data cleaning?") and get answers from the LLM.

### Fine-Tuning Data Export

To help you build custom LLMs tailored to your codebase or documentation style, `deepwikiR` can export the documented code elements. This feature is controlled by the `export_finetuning_data` section in each project's configuration:

```json
{
  // ... other project settings ...
  "export_finetuning_data": {
    "enabled": true,
    "output_filename": "my_project_finetune_data.jsonl",
    "format": "jsonl_chat" // or "jsonl_prompt_completion"
  }
}
```

*   **`enabled`**: Set to `true` to activate this feature.
*   **`output_filename`**: The name of the JSONL file to be created in the project's `output_dir`.
*   **`format`**:
    *   `"jsonl_chat"`: Exports data as a sequence of system, user (code context), and assistant (documentation) messages.
    *   `"jsonl_prompt_completion"`: Exports data as direct prompt (code context) and completion (documentation) pairs.

The generated JSONL file can then be used with various LLM fine-tuning platforms and libraries.

## 5. Usage

### Command-Line Interface (CLI)

The primary interface for `deepwikiR` is its command-line tool, invoked via `Rscript R/deepwikiR.R`.

**Generate Documentation:**
To process all projects defined in your configuration file:
```{bash eval=FALSE}
# Assuming you are in the root of the deepwikiR project directory
Rscript R/deepwikiR.R generate --config inst/config/sample_config.json
```
You can add `--quiet` to suppress verbose logging or `--api-key YOUR_KEY_HERE` to override the API key for all projects processed in that run.

**Chat with Codebase:**
(As shown in the "Interactive Chat Mode" section above).

### From an R Session

While the CLI is recommended for batch processing, you can also invoke `deepwikiR` functions from an R session, especially for development or custom workflows.

```{r eval=FALSE}
# Load all functions if working from source (or library(deepwikiR) if installed)
# devtools::load_all()

# 1. Load configuration
config_data <- deepwikiR::load_and_validate_config("inst/config/sample_config.json")

# 2. Process a specific project (e.g., the first one)
if (length(config_data$projects) > 0) {
  first_project_config <- config_data$projects[[1]]

  # Generate documentation for this project
  output_path <- deepwikiR::generate_repo_docs(
    project_config = first_project_config,
    verbose = TRUE
  )

  if (!is.null(output_path)) {
    cat("Documentation generated at:", output_path, "\n")

    # If you want to chat with this newly documented project:
    # Construct path to its .rds file
    rds_filename <- paste0(
      tools::file_path_sans_ext(first_project_config$output_filename_base),
      "_analysis_data.rds"
    )
    analysis_rds_path <- file.path(first_project_config$output_dir, rds_filename)

    if (file.exists(analysis_rds_path)) {
      deepwikiR::chat_with_repo(
        analysis_data_path = analysis_rds_path,
        config_path = "inst/config/sample_config.json" # Path to the main config
      )
    }
  }
}
```

## 6. Comprehensive Demo: Analyzing `r-lib/cli`

`deepwikiR` includes an example configuration and script to demonstrate its capabilities on a real-world R package, `r-lib/cli`.

**Important:** Running this demo involves:
*   Cloning the `r-lib/cli` repository (which is moderately sized).
*   Making numerous LLM API calls, which will consume tokens and may incur costs depending on your LLM provider.
*   Significant processing time.

**Therefore, this demo should be run manually and is not executed as part of the vignette build.**

### Demo Configuration (`inst/examples/r-lib-cli-config.json`)

The demo uses a dedicated configuration file. Key aspects:

```json
{
  "global_settings": {
    "output_dir_root": "output/deepwikiR_demos", // Demo outputs will be here
    "default_llm_settings": {
      "provider": "openai",
      "model": "gpt-3.5-turbo",
      // ... other LLM settings ...
      "api_details": {
        "api_key_env_var": "OPENAI_API_KEY"
      }
    }
  },
  "projects": [
    {
      "project_name": "r-lib-cli-demo",
      "code_source": {
        "git_repo": "https://github.com/r-lib/cli.git", // Target repository
        "language_hints": ["R"],
        "include_patterns": ["*.R"],
        "exclude_patterns": ["tests/testthat/helper-*.R", "tests/testthat/setup-*.R", "revdep/"]
      },
      "output_dir": "r-lib-cli-docs", // Subdirectory for this project's docs
      "output_filename_base": "rlib_cli_docs",
      "quarto_format": "html",
      "export_finetuning_data": { // Fine-tuning data will be exported
        "enabled": true,
        "output_filename": "rlib_cli_finetune_export.jsonl",
        "format": "jsonl_chat"
      }
    }
  ]
}
```
This configuration tells `deepwikiR` to:
1.  Clone `r-lib/cli` from GitHub.
2.  Process its R files, excluding specified test helpers and revdep patterns.
3.  Generate HTML documentation into `output/deepwikiR_demos/r-lib-cli-docs/`.
4.  Export fine-tuning data to `rlib_cli_finetune_export.jsonl` in the same output directory.

### Running the Demo

The `inst/examples/run_rlib_cli_demo.R` script is provided to execute this demo.

1.  **Ensure `deepwikiR` is usable:** Either install it or load its functions (e.g., `devtools::load_all()` if you are in the `deepwikiR` project root).
2.  **Set your LLM API Key:** Make sure the environment variable specified in `r-lib-cli-config.json` (e.g., `OPENAI_API_KEY`) is set.
3.  **Execute the script from the `deepwikiR` project root:**
    ```{bash eval=FALSE}
    Rscript inst/examples/run_rlib_cli_demo.R
    ```
    Alternatively, you can open and run this R script interactively from your R environment.

### Expected Outcome

*   The `r-lib/cli` repository will be cloned (first time only, or if the temporary clone is removed).
*   `deepwikiR` will process the R files, making calls to the configured LLM. This will take some time.
*   You will see verbose output in the console tracking the progress.
*   Upon completion, the following will be generated in `output/deepwikiR_demos/r-lib-cli-docs/`:
    *   `rlib_cli_docs.html` (the main Quarto documentation).
    *   `rlib_cli_docs_analysis_data.rds` (the data file for chat mode).
    *   `rlib_cli_finetune_export.jsonl` (the exported data for fine-tuning).
*   The script will also print a command you can use to start chatting with the `r-lib/cli` codebase:
    ```{bash eval=FALSE}
    # Example command that might be printed (paths will be absolute):
    Rscript R/deepwikiR.R chat \
      --data_file /path/to/your/deepwikiR/output/deepwikiR_demos/r-lib-cli-docs/rlib_cli_docs_analysis_data.rds \
      --config_file /path/to/your/deepwikiR/inst/examples/r-lib-cli-config.json
    ```

### Illustrative Output Snippet (Conceptual)

The generated `rlib_cli_docs.html` would contain sections like:

**Directory Overview**

*   **Directory: `R/`**
    This directory contains the core R source code for the `cli` package. It includes implementations for features like theming, string manipulation, progress bars, tree structures, and various UI elements for command-line interfaces...

**Detailed Code Element Documentation**

*   **`ansi_has_any()`**
    *   **Type:** function
    *   **Signature:** `ansi_has_any(x)`
    *   **File:** `R/ansi-common.R`
    *   **Description:**
        This function checks if any of the strings in the input character vector `x` contain ANSI escape sequences. It is useful for determining if a string has existing ANSI styling...
    *   **Code:**
        ```r
        ansi_has_any <- function(x) {
          grepl(ansi_regex(), x, perl = TRUE)
        }
        ```

## 7. Troubleshooting

*   **API Key Issues:** Ensure your LLM API key is correctly set as an environment variable and that the `api_key_env_var` in your config matches this variable name. Check for typos and ensure your API account has credit/access.
*   **Git Cloning Failures:** Verify the `git_repo` URL in your config. Ensure Git is installed and accessible in your system's PATH. For private repositories, SSH key setup might be necessary.
*   **Quarto Rendering Errors:** Make sure Quarto CLI is installed and in your PATH. Check the console output from `deepwikiR` for specific error messages from Quarto.
*   **Long Processing Times:** Documenting large repositories or using larger LLM models can take significant time and consume many tokens. Monitor progress via verbose output. Consider refining `include_patterns` and `exclude_patterns` in your config to process only essential parts of very large projects.

## 8. Conclusion

`deepwikiR` aims to significantly streamline the process of documenting and understanding R code repositories. By integrating static analysis with LLM-driven content generation and interactive exploration, it offers a powerful toolkit for R developers.

Future development may include deeper static analysis, more sophisticated call graph interactions, support for additional languages, and enhanced customization options.

Contributions and feedback are welcome!
